Text embedding models reveal high resolution insights into conceptual knowledge from short multiple choice quizzes

### 2023-01-26 paper/main.tex
flattened transparency in Fig 1, lots of minor wording tweaks, added a topic comparison figure + associated text to the results section

\title{Geometric models reveal the hidden structure of conceptual knowledge}

@@

\begin{abstract} We develop a mathematical framework, based on natural language
processing models, for tracking and characterizing the acquisition of
conceptual knowledge. Our approach embeds each concept in a high dimensional
representation space, where nearby coordinates reflect similar or related
concepts. We tested our approach using behavioral data collected from a group
of college students. In the experiment, we asked the participants to answer
sets of quiz questions interleaved between watching two course videos from the
Khan Academy platform. We applied our framework to the videos' transcripts, and
to text of the quiz questions, to quantify the content of each moment of video
and each quiz question. We used these embeddings, along with participants' quiz
responses, to track how the learners' knowledge changed after watching each
video. Our findings show how a limited set of quiz questions may be used to
construct rich and meaningful representations of what each learner knows, and
how their knowledge changes over time as they learn.

@@

Suppose that a teacher had access to a complete ``map'' of everything a student
knew. Defining what such a map might even look like, let alone how it might be
constructed or filled in, is itself a non-trivial problem. But if a teacher
\textit{were} to gain access to such a map, how might that change their ability
to teach the student? Perhaps they might start by checking how well the student
knew the to-be-learned information already, or how much they knew about related
concepts. For some students, they could potentially optimize their teaching
efforts to maximize efficiency by focusing primarily on not-yet-known content.
For other students (or other content areas), it might be more effective to
optimize for direct connections between already-known content and any new
material. Observing how the student's knowledge was changing over time, in
response to their training, could also help to guide the teacher.

@@

Here we develop a framework for modelling how knowledge is acquired during
learning. The central idea is to use text embedding models to define the
coordinate systems of two maps: (a) a \textit{knowledge map} that describes the
extent to which each concept is currently known and (b) a \textit{learning map}
that describes the extent to which each concept could be learned. Each location
on these maps represents a single concept, and the geometries are defined such
that related concepts are located nearby in space. We use this framework to
analyze and interpret behavioral data collected from an experiment that has
participants watch and answer conceptual questions about a series of recorded
course lectures.

@@

We asked our participants to answer questions from several multiple choice
quizzes and watch two lecture videos from the \textit{Khan Academy} platform
(Fig.~\ref{fig:experiment}). One lecture video, entitled \textit{Four
Fundamental Forces}, was about the four fundamental forces in physics: gravity,
strong and weak interactions, and electromagnetism. The second lecture video,
entitled \textit{Birth of Stars}, provides an overview of our current
understanding of how stars form. We selected both lessons to be (a) accessible
to a broad audience, e.g., by minimizing prerequisite knowledge, (b) largely
independent of each other, e.g., so that the two videos focused on different
material and did not depend on each other, and (c) related to each other, e.g.,
so that both videos contained at least \textit{some} similar or overlapping
content. The two videos we selected are introductory, about different primary
concepts, but also touch on ``physics'' and ``astronomy'' themes. We also wrote
a set of multiple choice quiz questions that would enable us to test
participants' knoweldge about each individual video and about related content
not specifically presented in either video (Tab.~\questions). Participants
answered questions randomly drawn from each content area (lecture 1, lecture 2,
and general physics knowledge) across each of three quizzes. Quiz 1 was
intended to assessed participants' knowledge before training; quiz 2 assessed
knowledge after watching the Four Fundamental Forces video (i.e., lecture 1);
and quiz 3 assessed knowledge after watching the Birth of Stars video (i.e.,
lecture 2).

@@

We trained a text embedding model using sliding windows of text from the two
videos' transcripts (see \textit{Constructing text embeddings of multiple
videos and questions}). We also used the same model (i.e., trained on the
videos' transcripts) to embed the text of each question in our pool. This
yielded, for each second of each video, and for each question, a single topic
vector-- i.e., a coordinate in a text embedding space
(Fig.~\ref{fig:sliding-windows}). Intuitively, each dimension of the embedding
space corresponds to a ``theme'' or ``topic'' reflected in some part(s) of the
videos (Tab.~\topics), and the coordinates in embedding space denote the blend
of themes reflected by a particular excerpt of text (e.g., from part of a
video's transcript, from a question, etc.).

@@

    \caption{\textbf{Which parts of each lecture are captured by each
    question?} Each panel displays timeseries plots showing how each question's
    topic vector correlates with each video timepoint's topic vector. The left
    panel displays these correlations for the \textit{Four Fundamental Forces}
    lecture and associated questions, and the right panels displays these
    correlations for the \textit{Birth of Stars} lecture and associated
    questions. The colors denote question identities. The diamonds in each
    panel denote the moment of peak correlation between the indicated
    questions, in the indicated lectures. The associated questions' text, and
    snippets of the lectures' transcripts in the best-matching sliding windows,
    are displayed at the bottom of the figure.}

@@

Although a single lecture may be organized a single broad theme at a coarse
scale, at a finer scale each moment of a lecture typically covers a narrower
range of content. We wondered whether a text embedding model trained on the
lectures' transcripts might capture some of this finer scale content. For
example, if a particular question asks about the content from one small part of
a lecture, we wondered whether our text embedding model could be used to
automatically identify the ``matching'' moment(s) in the lecture. When we
correlated each question's topic vector with the topic vectors for each second
of the lectures, we found some evidence that each question is temporally
specific (Fig.~\ref{fig:question-correlations}). In particular, most questions'
topic vectors were maximally correlated with a well-defined range of timepoints
from their corresponding lectures, and the correlations fell off sharply
outside of that range. We also examined the best-matching intervals for each
question qualitatively by comparing the text of the question to the text of the
most-correlated parts of the lectures. Despite that the questions were excluded
from the text embedding model's training set, in general we found a close
correspondence between the conceptual content that each question covered and
the content covered by the best-matching moments of the lectures. Two
representative examples are shown at the bottom of
Fig.~\ref{fig:question-correlations}.

The ability to quantify how much each question is ``asking about'' each moment
of the lectures could enable high-resolution insights into participants'
knowledge. Traditional approaches to estimating how much a student ``knows''
about the content of a given lecture entail computing the proportion of
correctly answered questions. But if two students receive identical scores on
an exam, might our modeling framework help us to gain more naunced insights
into the \textit{specific} content that each student has mastered (or failed to
master)? For example, a student who misses three questions that were all about
the same concept (e.g., concept $A$) will have gotten the same
\textit{proportion} of questions correct as another student who missed three
questions about three \textit{different} concepts (e.g., $A$, $B$, and $C$).
But if we wanted to fill in the ``gaps'' in the two students' understandings,
we might do well to focus on concept $A$ for the first student, but to also add
in materials pertaining to concepts $B$ and $C$ for the second student.

We developed a simple formula (Eqn.~\ref{eqn:prop}) for using a participant's
responses to a small set of multiple choice questions to estimate how much the
participant ``knows'' about the concept reflected by any arbitrary coordinate,
$x$, in text embedding space (e.g., the content reflected by any moment in a
lecture they had watched; see \textit{Estimating dynamic knowledge traces}).
Essentially, the estimated knowledge at the coordinate is given by the weighted
average proportion of quiz questions the participant answered correctly, where
the weights reflect how much each question is ``about'' the content at $x$.
When we apply this approach to estimate the participant's knowledge about the
content presented in each moment of each lecture, we can obtain a detailed
timecourse describing how much ``knowledge'' the participant has about any part
of the lecture. As shown in Figures~\ref{fig:knowledge-timeseries}A and C), we
can also apply this approach separately for the questions from each quiz the
participants took throughout the experiment. From just 13 questions per quiz,
we obtain a high-resolution snapshot (at the time each quiz was taken) of what
the participants knew about any moment's content, from either of the two
lectures they watched (comprising a total of 1106 samples across the two
lectures).

@@

If we are able to accurately estimate a participant's knowledge about the
content tested by a given question, the estimated knowledge should have some
predictive information about whether the participant is likely ot answer the
question correctly or incorrectly. For each question in turn, for each
participant, we used Equation~\ref{eqn:prop} to estimate (using all
\textit{other} questions from the same quiz, from the same participant) the
participant's knowledge at the held-out question's embedding coordinate. For
each quiz, we aggregated these estimates into two distributions: one for the
estimated knowledge at the coordinates of each \textit{correctly} answered
question, and another for the estimated knowledge at the coordinates of each
\textit{incorrectly} answered question (Fig.~\ref{fig:predictions}). We then
used independent samples $t$-tests to compare the means of these distributions
of estimated knowledge.

@@

Knowledge estimates need not be limited to the content of the lectures. As
illustrated in Figure~\ref{fig:knowledge-maps}, our general approach to
estimating knowledge from a small number of quiz questions may be applied to
\textit{any} content, given its text embedding coordinate. To visualize how
knowledge ``spreads'' through text embedding space to content beyond the
lectures participants watched, we first fit a new topic model to the lectures'
sliding windows with $k = 100$ topics. We hoped that increasing the number of
topics from 15 to 100 might help us to generalize the knowledge predictions.
(Aside from increasing the number of topics from 15 to 100, all other
procedures and model parameters were carried over from the preceding analyses.)
We resampled each lecture's topic trajectory to 1~Hz and also projected each
question into the same 100-dimensional topic space.

@@

    \caption{\textbf{Mapping out the geometry of knowledge and learning.}
    \textbf{A. Average ``knowledge maps'' estimated using each quiz.} Each map
    displays a 2D projection of the estimated knowledge about the content
    reflected by \textit{all} regions of topic space (see \textit{Creating
    knowledge and learning map visualizations}). The topic trajectories of each
    lecture and the coordinates of each question are indicated by dotted lines
    and dots. Each map reflects an average across all participants. For
    individual participants' maps, see
    Figures~\individualKnowledgeMapsA,~\individualKnowledgeMapsB,
    and~\individualKnowledgeMapsC. \textbf{B. Average ``learning maps''
    estiamted between each successive pair of quizzes.} The learning maps are
    in the same general format as the knowledge maps in Panel A, but each
    coordinate in the learning maps indicates the \textit{difference} between
    the corresponding coordinates in the indicated \textit{pair} of knowledge
    maps-- i.e., how much the estimated knowledge ``changed'' across the two
    quizzes. Each map reflects an average across all participants. For
    individual participants' maps, see
    Figures~\individualLearningMapsA~and~\individualLearningMapsB. \textbf{C.
    Word clouds for sampled points in topic space.} Each word cloud displays
    the relative weights of each word reflected by the blend of topics
    represented at the locations of the stars in the maps. The words' colors
    indicate how much each word is weighted on average across all timepoints'
    topic vectors in the \textit{Four Fundamental Forces} (blue) and
    \textit{Birth of Stars} (green) videos, respectively.}

@@

We projected the resulting 100-dimensional topic vectors (for each second of
video and for each question) into a shared 2-dimensional space (see
\textit{Creating knowledge and learning map visualizations}). Next, we sampled
points evenly from a $100 \times 100$ grid of coordinates that evenly tiled a
rectangle enclosing the 2D projections of the video trajectories and questions.
We used Equation~\ref{eqn:rbf-knowledge} to estimate participants' knowledge at
each of these 10K sampled locations, and we averaged these estimates across
participants to obtain an estimated average \textit{knowledge map}
(Fig.~\ref{fig:knowledge-maps}). Intuitively, the knowledge map constructed
from a given quiz's responses provides a visualization of how ``much''
participants know about any content expressible by the fitted model.

Several features of the resulting knowledge maps are worth noting. The average
knowledge map estimated from Quiz 1 responses (Fig.~\ref{fig:knowledge-maps},
leftmost map) shows that participants tended to have relatively little
knowledge about any parts of the text embedding space (i.e., the shading is
relatively dark everywhere). The knowledge map estimated from Quiz 2 responses
shows a marked increase in knowledge on the left side of the map (around
roughly the same range of coordinates covered by the \textit{Four Fundamental
Forces} lecture, indicated by the dotted blue line). In other words,
participants' estimated increase in knowledge is localized to conceptual
content that is nearby (i.e., related to) the content from the lecture they
watched prior to taking Quiz 2. This localization is non-trivial: the knowledge
estimates are informed only by the embedded coordinates of the \textit{quiz
questions}, not by the embeddings of either lecture. Finally, the knowledge map
estimated from Quiz 3 responses shows a second increase in knowledge, localized
to the region surrounding the embedding of the \textit{Birth of Stars} lecture
participants watched prior to taken Quiz 3.

@@

Because the 2D projection we used to construct the knowledge and learning maps
is (partially) invertable, we may gain additional insights into the estimates
by reconstructing the original high-dimensional topic vectors for any point(s)
in the maps we are interested in. For example, this could serve as a useful
tool for an instructor looking to better understand which content areas a
student (or a group of students) knows well (or poorly). As a demonstration, we
show the top-weighted words from the blends of topics reconstructed from three
locations on the maps (Fig.~\ref{fig:knowledge-maps}C): one point near the
\textit{Four Fundamental Forces} embedding (yellow); a second point near the
\textit{Birth of Stars} embedding (orange), and a third point somewhere in
between the two lectures' embeddings (pink). As shown in the word clouds in the
Panel, the top-weighted words at the coordinate near the \textit{Four
Fundamental Forces} embedding also tended to be weighted heavily by the topics
expressed in that lecture. Similarly, the top-weighted words at the coordinate
near the \textit{Birth of Stars} embedding tended to be weighted most heavily
by the topics expressed in \textit{that} lecture. And the top-weighted words at
the coordinate between the two lectures' embeddings show a roughly even mix of
words most strongly associated with each lecture.

@@

A total of 49 participants reporting having normal hearing and 1
participant reported having some hearing impairment. A total of 49
participants reported having normal color vision and 1 participant
reported being color blind.  Participants reported having had, on the
night prior to testing, 2 -- 4 hours of sleep (1 participant), 4 -- 6
hours of sleep (9 participants), 6 -- 8 hours of sleep (35
participants), or $8+$ hours of sleep (5 participants). They reported
having consumed, on the same day and leading up to their testing
session, 0 cups of coffee (38 participants), 1 cup of coffee (10
participants), 3 cups of coffee (1 participant), or $4+$ cups of
coffee (1 participant).

@@

Particpants reported their undergraduate major(s) as Social Sciences
(28 participants), Natural sciences (16), Professional (e.g., pre-med
or pre-law; 8 participants), Mathematics and engineering (7
participants), Humanities (4 participants), or Undecided (3
participants).  Note that some participants selected multiple
categories for their undergraduate major.  We also asked participants
about the courses they had taken.  In total, 46 participants reported
having taken at least one Khan academy course in the past or being
familiar with the Khan academy, and 4 reported not having taken any
Khan academy courses.  Of the participants who reported having watched
at least one Khan academy course, 1 participant declined to report the
number of courses they had watched; 7 participants reported having
watched 1--2 courses; 11 reported having watched 3--5 courses; 8
reported having watched 5--10 courses; and 19 reported having watched
10 or more courses.  We also asked participants about the specific
courses they had watched, categorized under different subject areas.
In the ``Mathematics'' area participants reported having watched
videos on AP Calculus AB (21 participants), Precalculus (17
participants), Algebra 2 (14 participants), AP Calculus BC (12
participants), Trigonometry (11 participants), Algebra 1 (10
participants), Geometry (8 participants), Pre-algebra (7
participants), Multivariable Calculus (5 participants), Differential
Equations (5 participants, Statistics and Probability (4
participants), AP Statistics (2 participants), Linear Algebra (2
participants), Early Math (1 participant), Arithmetic (1 participant),
and other videos not listed in our survey (6 participants).  In the ``Science and engineering''
area participants reported having watched videos on Chemistry, AP
Chemistry, or Organic Chemstry (21
participants); Physics, AP Physics I, or AP Physics II (15 participants); Biology, AP
Biology; or High school Biology (15 participants); Health and Medicine
(1 participant); or other videos not listed in our survey (20 participants).  We also asked
participants if they had specifically seen the videos used in our
experiment.  When we asked about the \textit{Four Fundamental Forces}
video, 45 participants reported not having watched it before, 1
participant reported that they were not sure if they had watched it
before, and 4 participants declined to respond.  When we asked about
the \textit{Birth of Stars} video, 46 participants reported not having
watched it before and 4 participants declined to respond.  When we
asked participants about non-Khan academy online courses, they
reported having watched or taken courses on Mathematics (15
participants), Science and engineering (11 participants), Test
preparation (9 participants), Economics and finance (3 participants),
Arts and humanities (2 participants), Computing (2 participants), and
other categories not listed in our survey (18 participants).  Finally,
we asked participants about in-person courses they had taken in
different subject areas.  They reported taking courses in Mathematics
(39 participants), Science and engineering (38 participants), Arts and
humanities (35 participants), Test preparation (27 participants),
Economics and finance (26 participants), Computing (15 participants),
College and careers (7 participants), or other courses not listed in
our survey (6 participants).

@@

After fitting a topic model to each videos' transcripts, we could use the
trained model to transform arbitrary (potentially new) documents into
$k$-dimensional topic vectors. A convenient property of these topic vectors is
that documents that reflect similar blends of topics (i.e., documents that
reflect similar themes, according to the model) will yield similar (in terms of
Euclidean distance, correlation, etc.) topic vectors. In general, the
similarity between different documents' topic vectors may be used to
characterize the similarity in content between the documents.

>>>>

\title{Text embedding models reveal high resolution insights into conceptual
knowledge from short multiple choice quizzes}

@@

\begin{abstract} We develop a mathematical framework, based on natural language
processing models, for tracking and characterizing the acquisition of
conceptual knowledge. Our approach embeds each concept in a high dimensional
representation space, where nearby coordinates reflect similar or related
concepts. We test our approach using behavioral data collected from a group of
college students. In the experiment, we ask the participants to answer small
sets of multiple choice quiz questions interleaved between watching two course
videos from the Khan Academy platform. We applied our framework to the videos'
transcripts, and to text of the quiz questions, to quantify the content of each
moment of video and each quiz question. We used these embeddings, along with
participants' quiz responses, to track how the learners' knowledge changed
after watching each video. Our findings show how a small set of quiz questions
may be used to obtain rich and meaningful high resolution insights into what
each learner knows, and how their knowledge changes over time as they learn.

@@

Suppose that a teacher had access to a complete ``map'' of everything their
student knew. Defining what such a map might even look like, let alone how it
might be constructed or filled in, is itself a non-trivial problem. But if a
teacher \textit{were} to gain access to such a map, how might that change their
ability to teach the student? Perhaps they might start by checking how well the
student knew the to-be-learned information already, or how much they knew about
related concepts. For some students, they could potentially optimize their
teaching efforts to maximize efficiency by focusing primarily on not-yet-known
content. For other students (or other content areas), it might be more
effective to optimize for direct connections between already-known content and
any new material. Observing how the student's knowledge was changing over time,
in response to their training, could also help to guide the teacher.

@@

Here we develop a framework for modelling how knowledge is acquired during
learning. The central idea is to use text embedding models to define the
coordinate systems of two maps: (a) a \textit{knowledge map} that describes the
extent to which each concept is currently known and (b) a \textit{learning map}
that describes changes in knowledge over time. Each location on these maps
represents a single concept, and the maps' geometries are defined such that
related concepts are located nearby in space. We use this framework to analyze
and interpret behavioral data collected from an experiment that has
participants watch and answer multiple choice questions about a series of
recorded course lectures.

@@

@@

We asked our participants to answer questions from several multiple choice
quizzes and watch two lecture videos from the \textit{Khan Academy} platform
(Fig.~\ref{fig:experiment}). One lecture video, entitled \textit{Four
Fundamental Forces}, was about the four fundamental forces in physics: gravity,
strong and weak interactions, and electromagnetism. The second lecture video,
entitled \textit{Birth of Stars}, provides an overview of our current
understanding of how stars form. We selected both lessons to be (a) accessible
to a broad audience, e.g., by minimizing prerequisite knowledge, (b) largely
independent of each other, e.g., so that the two videos focused on different
material and did not depend on each other, and (c) related to each other, e.g.,
so that both videos contained at least \textit{some} similar or overlapping
content. The two videos we selected are introductory (i.e., minimizing specific
prerequisite knowledge), and are about different primary concepts, but they
both also touch on ``physics'' and ``astronomy'' themes. We also wrote a set of
multiple choice quiz questions that would enable us to test participants'
knoweldge about each individual video and about related content not
specifically presented in either video (Tab.~\questions). Participants answered
questions randomly drawn from each content area (lecture 1, lecture 2, and
general physics knowledge) across each of three quizzes. Quiz 1 was intended to
assessed participants' knowledge before training; quiz 2 assessed knowledge
after watching the \textit{Four Fundamental Forces} video (i.e., lecture 1);
and quiz 3 assessed knowledge after watching the \textit{Birth of Stars} video
(i.e., lecture 2).

We trained a text embedding model using sliding windows of text from the two
videos' transcripts (see \textit{Constructing text embeddings of multiple
videos and questions}). We also used the same model (i.e., trained on the
videos' transcripts) to embed the text of each question in our pool. This
yielded, for each second of each video, and for each question, a single
15-dimensional topic vector---i.e., a coordinate in a text embedding space
(Fig.~\ref{fig:sliding-windows}). Intuitively, each dimension of the embedding
space corresponds to a ``theme'' or ``topic'' reflected in some part(s) of the
videos (Tab.~\topics), and the coordinates in embedding space denote the blend
of themes reflected by a particular excerpt of text (e.g., from part of a
video's transcript, from a question, etc.).

\begin{figure}[tp]
    \centering
    \includegraphics[width=0.6\textwidth]{figs/active-topics}

\caption{\textbf{Lecture and question topic overlap.} The bar plots display the
variability in topic weights across lecture timepoints (top panels) and
questions (bottom panels); colors denote topics. The top-weighted words from
the most ``expressive'' (i.e., variable across observations) topics from each
lecture are displayed in the upper right (orange: topic 2; yellow-green: topic
4). The top-weighted words from the full set of topics may be found in
Table~\topics.}

    \label{fig:topics}
\end{figure}

Given that we trained the text embedding model using the video transcripts, we
wondered whether the questions that were (ostensibly, by design) ``about'' the
content of each lecture would ``match up'' correctly with the lectures. In
other words, we hoped that the text embeddings would capture something about
the deeper conceptual content of the lectures, beyond surface details such as
exact wording choices. If so, when we embed \textit{new} text outside of the
model's training set, we should see a correspondance between the embeddings of
the training data (i.e., snippets of text from the lectures' transcripts) and
other text that reflects related concepts (e.g., questions \textit{about} each
lecture). Further, although the content from any given moment from a lecture
might stray from the average content (across all timepoints), we hoped that
\textit{variability} in each topic's expression over timepoints within a
lecture would match up with the variability in topic expressions for questions
about that lecture. Intuitively, the variability in the expression of a given
topic relates to how much ``information''~\citep{Fish22} the lecture (or
questions) reflect about that topic. When we compared the variability in topic
weights across each lecture's timepoints with the variability in topic weights
across each question set, we found a strong correspondence
(Fig.~|ref{fig:topics}). The most variable topics from the \textit{Four
Fundamental Forces} lecture, and questions about that lecture, are 2, 3, 8, 12,
13, 14, and 15. The most variable topics from the \textbf{Birth of Stars}
lecture, and questions about that lecture, are 1, 4, 5, 6, and 7. This strong
overlap between the lectures and questions specifically about each lecture
indicates that the topic model captures some of the underlying conceptual
content.

@@

    \caption{\textbf{Which parts of each lecture are captured by each
    question?} Each panel displays timeseries plots showing how each question's
    topic vector correlates with each video timepoint's topic vector (Panel
    \textbf{A.}: correlations for the \textit{Four Fundamental Forces} lecture
    and associated questions; Panel \textbf{B.}: correlations for the
    \textit{Birth of Stars} lecture and associated questions). The colors denote
    question identities. The diamonds in each panel denote the moment of peak
    correlation between the indicated questions, in the indicated lectures. The
    associated questions' text, and snippets of the lectures' transcripts in
    the best-matching sliding windows, are displayed at the bottom of the
    figure.}

@@

Although a single lecture may be organized around a single broad theme at a
coarse scale, at a finer scale each moment of a lecture typically covers a
narrower range of content. We wondered whether a text embedding model trained
on the lectures' transcripts might capture some of this finer scale content.
For example, if a particular question asks about the content from one small
part of a lecture, we wondered whether our text embedding model could be used
to automatically identify the ``matching'' moment(s) in the lecture. When we
correlated each question's topic vector with the topic vectors for each second
of the lectures, we found some evidence that each question is temporally
specific (Fig.~\ref{fig:question-correlations}). In particular, most questions'
topic vectors were maximally correlated with a well-defined (and relatively
narrow) range of timepoints from their corresponding lectures, and the
correlations fell off sharply outside of that range. We also examined the
best-matching intervals for each question qualitatively by comparing the text
of the question to the text of the most-correlated parts of the lectures.
Despite that the questions were excluded from the text embedding model's
training set, in general we found (through manual inspection) a close
correspondence between the conceptual content that each question covered and
the content covered by the best-matching moments of the lectures. Two
representative examples are shown at the bottom of
Fig.~\ref{fig:question-correlations}.

The ability to quantify how much each question is ``asking about'' the content
from each moment of the lectures could enable high-resolution insights into
participants' knowledge. Traditional approaches to estimating how much a
student ``knows'' about the content of a given lecture entail computing the
proportion of correctly answered questions. But if two students receive
identical scores on an exam, might our modeling framework help us to gain more
naunced insights into the \textit{specific} content that each student has
mastered (or failed to master)? For example, a student who misses three
questions that were all about the same concept (e.g., concept $A$) will have
gotten the same \textit{proportion} of questions correct as another student who
missed three questions about three \textit{different} concepts (e.g., $A$, $B$,
and $C$). But if we wanted to fill in the ``gaps'' in the two students'
understandings, we might do well to focus on concept $A$ for the first student,
but to also add in materials pertaining to concepts $B$ and $C$ for the second
student.

We developed a simple formula (Eqn.~\ref{eqn:prop}) for using a participant's
responses to a small set of multiple choice questions to estimate how much the
participant ``knows'' about the concept reflected by any arbitrary coordinate,
$x$, in text embedding space (e.g., the content reflected by any moment in a
lecture they had watched; see \textit{Estimating dynamic knowledge traces}).
Essentially, the estimated knowledge at the coordinate is given by the weighted
average proportion of quiz questions the participant answered correctly, where
the weights reflect how much each question is ``about'' the content at $x$.
When we apply this approach to estimate the participant's knowledge about the
content presented in each moment of each lecture, we can obtain a detailed
timecourse describing how much ``knowledge'' the participant has about any part
of the lecture. As shown in Figure~\ref{fig:knowledge-timeseries}, we can also
apply this approach separately for the questions from each quiz the
participants took throughout the experiment. From just 13 questions per quiz,
we obtain a high-resolution snapshot (at the time each quiz was taken) of what
the participants knew about any moment's content, from either of the two
lectures they watched (comprising a total of 1106 samples across the two
lectures).

@@

If we are able to accurately estimate a participant's knowledge about the
content tested by a given question, the estimated knowledge should have some
predictive information about whether the participant is likely to answer the
question correctly or incorrectly. For each question in turn, for each
participant, we used Equation~\ref{eqn:prop} to estimate (using all
\textit{other} questions from the same quiz, from the same participant) the
participant's knowledge at the held-out question's embedding coordinate. For
each quiz, we aggregated these estimates into two distributions: one for the
estimated knowledge at the coordinates of each \textit{correctly} answered
question, and another for the estimated knowledge at the coordinates of each
\textit{incorrectly} answered question (Fig.~\ref{fig:predictions}). We then
used independent samples $t$-tests to compare the means of these distributions
of estimated knowledge.

@@

Knowledge estimates need not be limited to the content of the lectures. As
illustrated in Figure~\ref{fig:knowledge-maps}, our general approach to
estimating knowledge from a small number of quiz questions may be applied to
\textit{any} content, given its text embedding coordinate. To visualize how
knowledge ``spreads'' through text embedding space to content beyond the
lectures participants watched, we first fit a new topic model to the lectures'
sliding windows with $k = 100$ topics. We hoped that increasing the number of
topics from 15 to 100 might help us to generalize the knowledge predictions.
(Aside from increasing the number of topics from 15 to 100, all other
procedures and model parameters were carried over from the preceding analyses.)
As in our other analyses, we resampled each lecture's topic trajectory to 1~Hz
and also projected each question into a shared text embedding space.

@@

    \caption{\textbf{Mapping out the geometry of knowledge and learning.}
    \textbf{A. Average ``knowledge maps'' estimated using each quiz.} Each map
    displays a 2D projection of the estimated knowledge about the content
    reflected by \textit{all} regions of topic space (see \textit{Creating
    knowledge and learning map visualizations}). The topic trajectories of each
    lecture and the coordinates of each question are indicated by dotted lines
    and dots. Each map reflects an average across all participants. For
    individual participants' maps, see
    Figures~\individualKnowledgeMapsA,~\individualKnowledgeMapsB,
    and~\individualKnowledgeMapsC. \textbf{B. Average ``learning maps''
    estiamted between each successive pair of quizzes.} The learning maps are
    in the same general format as the knowledge maps in Panel A, but each
    coordinate in the learning maps indicates the \textit{difference} between
    the corresponding coordinates in the indicated \textit{pair} of knowledge
    maps---i.e., how much the estimated knowledge ``changed'' across the two
    quizzes. Each map reflects an average across all participants. For
    individual participants' maps, see
    Figures~\individualLearningMapsA~and~\individualLearningMapsB. \textbf{C.
    Word clouds for sampled points in topic space.} Each word cloud displays
    the relative weights of each word reflected by the blend of topics
    represented at the locations of the stars in the maps. The words' colors
    indicate how much each word is weighted on average across all timepoints'
    topic vectors in the \textit{Four Fundamental Forces} (blue) and
    \textit{Birth of Stars} (green) videos, respectively.}

@@

We projected the resulting 100-dimensional topic vectors (for each second of
video and for each question) into a shared 2-dimensional space (see
\textit{Creating knowledge and learning map visualizations}). Next, we sampled
points evenly from a $100 \times 100$ grid of coordinates that evenly tiled a
rectangle enclosing the 2D projections of the videos and questions. We used
Equation~\ref{eqn:rbf-knowledge} to estimate participants' knowledge at each of
these 10K sampled locations, and we averaged these estimates across
participants to obtain an estimated average \textit{knowledge map}
(Fig.~\ref{fig:knowledge-maps}). Intuitively, the knowledge map constructed
from a given quiz's responses provides a visualization of how ``much''
participants know about any content expressible by the fitted text embedding
model.

Several features of the resulting knowledge maps are worth noting. The average
knowledge map estimated from Quiz 1 responses (Fig.~\ref{fig:knowledge-maps},
leftmost map) shows that participants tended to have relatively little
knowledge about any parts of the text embedding space (i.e., the shading is
relatively dark everywhere). The knowledge map estimated from Quiz 2 responses
shows a marked increase in knowledge on the left side of the map (around
roughly the same range of coordinates covered by the \textit{Four Fundamental
Forces} lecture, indicated by the dotted blue line). In other words,
participants' estimated increase in knowledge is localized to conceptual
content that is nearby (i.e., related to) the content from the lecture they
watched prior to taking Quiz 2. This localization is non-trivial: the knowledge
estimates are informed only by the embedded coordinates of the \textit{quiz
questions}, not by the embeddings of either lecture
(Eqn.~\ref{eqn:rbf-knowledge}). Finally, the knowledge map estimated from Quiz
3 responses shows a second increase in knowledge, localized to the region
surrounding the embedding of the \textit{Birth of Stars} lecture participants
watched prior to taking Quiz 3.

@@

Because the 2D projection we used to construct the knowledge and learning maps
is (partially) invertable, we may gain additional insights into the estimates
by reconstructing the original high-dimensional topic vectors for any point(s)
in the maps we are interested in. For example, this could serve as a useful
tool for an instructor looking to better understand which content areas a
student (or a group of students) knows well (or poorly). As a demonstration, we
show the top-weighted words from the blends of topics reconstructed from three
example locations on the maps (Fig.~\ref{fig:knowledge-maps}C): one point near
the \textit{Four Fundamental Forces} embedding (yellow); a second point near
the \textit{Birth of Stars} embedding (orange), and a third point somewhere in
between the two lectures' embeddings (pink). As shown in the word clouds in the
Panel, the top-weighted words at the example coordinate near the \textit{Four
Fundamental Forces} embedding also tended to be weighted heavily by the topics
expressed in that lecture. Similarly, the top-weighted words at the example
coordinate near the \textit{Birth of Stars} embedding tended to be weighted
most heavily by the topics expressed in \textit{that} lecture. And the
top-weighted words at the example coordinate between the two lectures'
embeddings show a roughly even mix of words most strongly associated with each
lecture.

@@

A total of 49 participants reporting having normal hearing and 1
participant reported having some hearing impairment. A total of 49
participants reported having normal color vision and 1 participant
reported being color blind.  Participants reported having had, on the
night prior to testing, 2--4 hours of sleep (1 participant), 4--6
hours of sleep (9 participants), 6--8 hours of sleep (35
participants), or $8+$ hours of sleep (5 participants). They reported
having consumed, on the same day and leading up to their testing
session, 0 cups of coffee (38 participants), 1 cup of coffee (10
participants), 3 cups of coffee (1 participant), or $4+$ cups of
coffee (1 participant).

@@

Particpants reported their undergraduate major(s) as Social Sciences (28
participants), Natural sciences (16), Professional (e.g., pre-med or pre-law; 8
participants), Mathematics and engineering (7 participants), Humanities (4
participants), or Undecided (3 participants). Note that some participants
selected multiple categories for their undergraduate major. We also asked
participants about the courses they had taken. In total, 46 participants
reported having taken at least one Khan academy course in the past or being
familiar with the Khan academy, and 4 reported not having taken any Khan
academy courses. Of the participants who reported having watched at least one
Khan academy course, 1 participant declined to report the number of courses
they had watched; 7 participants reported having watched 1--2 courses; 11
reported having watched 3--5 courses; 8 reported having watched 5--10 courses;
and 19 reported having watched 10 or more courses. We also asked participants
about the specific courses they had watched, categorized under different
subject areas. In the ``Mathematics'' area participants reported having watched
videos on AP Calculus AB (21 participants), Precalculus (17 participants),
Algebra 2 (14 participants), AP Calculus BC (12 participants), Trigonometry (11
participants), Algebra 1 (10 participants), Geometry (8 participants),
Pre-algebra (7 participants), Multivariable Calculus (5 participants),
Differential Equations (5 participants), Statistics and Probability (4
participants), AP Statistics (2 participants), Linear Algebra (2 participants),
Early Math (1 participant), Arithmetic (1 participant), and other videos not
listed in our survey (6 participants). In the ``Science and engineering'' area
participants reported having watched videos on Chemistry, AP Chemistry, or
Organic Chemstry (21 participants); Physics, AP Physics I, or AP Physics II (15
participants); Biology, AP Biology; or High school Biology (15 participants);
Health and Medicine (1 participant); or other videos not listed in our survey
(20 participants). We also asked participants if they had specifically seen the
videos used in our experiment. When we asked about the \textit{Four Fundamental
Forces} video, 45 participants reported not having watched it before, 1
participant reported that they were not sure if they had watched it before, and
4 participants declined to respond. When we asked about the \textit{Birth of
Stars} video, 46 participants reported not having watched it before and 4
participants declined to respond. When we asked participants about non-Khan
academy online courses, they reported having watched or taken courses on
Mathematics (15 participants), Science and engineering (11 participants), Test
preparation (9 participants), Economics and finance (3 participants), Arts and
humanities (2 participants), Computing (2 participants), and other categories
not listed in our survey (18 participants). Finally, we asked participants
about in-person courses they had taken in different subject areas. They
reported taking courses in Mathematics (39 participants), Science and
engineering (38 participants), Arts and humanities (35 participants), Test
preparation (27 participants), Economics and finance (26 participants),
Computing (15 participants), College and careers (7 participants), or other
courses not listed in our survey (6 participants).

@@

After fitting a topic model to each videos' transcripts, we could use the
trained model to transform arbitrary (potentially new) documents into
$k$-dimensional topic vectors. A convenient property of these topic vectors is
that documents that reflect similar blends of topics (i.e., documents that
reflect similar themes, according to the model) will yield similar (in terms of
Euclidean distance, correlation, etc.) topic vectors. In general, the
similarity between different documents' topic vectors may be used to
characterize the similarity in conceptual content between the documents.
